Fitness tracking reveals task-specific associations between memory, mental health, and exercise

### 2021-10-16 paper/main.tex
updated correlation figures and stats using 10000 bootstrap iterations

We used the same bootstrap procedure described above (used in our
exploratory correlation analyses) to generate $N = 1000$ bootstrapped
data matrices whose rows reflected sampled participants and whose
columns reflected different measured variables.

We grouped variables according to whether they were memory-related,
fitness-related, or demographic-related.  For each bootstrap
iteration, we divided the rows of that iterations data matrix into
training and test sets.  The assignments of rows to these two sets was
random, subject to the constraint that any duplicated rows in the data
matrix (i.e., reflecting a single participant who had been sampled
multiple times) was always assigned to either the training \textit{or}
the test set--i.e., duplicated rows could not appear in both the
training and the test sets.  The training sets always comprised 75\%
of the data, and the tests sets comprised the remaining 25\% of the
data.

Next, we fit a series of ridge regression models to the training data.
Specifically, for each pairing of memory, fitness, and demographic
variables, we fit a single ridge regression model treating the first
variable group as the input features and the second variable group as
the target features.  For example, one regression model used memory
variables to predict fitness variables, and another regression model
used fitness variables to predict demographic variables, and so on.
In total we fit six regression models to each training dataset.  We
then applied the fitted models to the held-out test dataset and
computed the root mean squared deviation (RMSD) between the predicted
and observed values in the target features of the test dataset.  We
also examined the regression weights assigned to each input feature.
This yielded, for each regression model (across $N$ bootstrap
iterations) a distribution of RMSD values and a distribution of
weights for each input variable.

We constructed a ``null'' distribution by using the same procedure as
above, but where the columns in the test datasets were randomly
permuted with each iteration (thereby breaking any meaningful
predictive information between the training and test data).  We
assessed the statistical significance ($p$-values) of the observed
RMSD values by computing the proportions of null RMSD values that were
less than the observed value.  We also assessed the significance of
the observed regression weights using $t$-tests to compare the means
of the observed versus null distributions of weights.

@@

\begin{itemize}
% \item characterizing behaviors (color by quartile and continue the
%   color scheme in later figures-- hue reflects task, shading reflects
%   performance.  white outline means immediate, black outline means delayed)
%   \begin{itemize}
%   \item Free recall (immediate + delayed): pfr, lag-CRP, spc  (color:
%     recall performance).  
%     Figure~\ref{fig:fr_behavioral}.
%     \item Naturalistic recall (immediate + delayed): reproduce a version of the sherlock
%       movie/recall trajectories (color: mean precision).  Figure~\ref{fig:nat_behavioral}.
%     \item Foreign language flashcards (immediate + delayed):
%       p(correct) histogram (color: p(correct)).  Figure~\ref{fig:vocab_behavioral}.
%       \item Spatial learning: mean error by number of shapes (color:
%         intercept and slope of line fit to errors as a function of the number of
%         shapes).  Figure~\ref{fig:spatial_behavioral}.
%       \end{itemize}
%     \item Fitness info (break down by task performance, potentially
%       separately for each task); also separate out recent (raw) and
%       recent versus baseline -- color using same color scheme as
%       behavior figure
%       \begin{itemize}
%       \item activity (steps, zone minutes, floors/elevation)
%       \item resting heart rate
%         \item sleep
%         \end{itemize}
  \item exploratory analysis (correlations).  Possibly make some sort of scatter matrix or pairplot-- rows/columns:
          tasks.  Diagonal: histogram of performance metric for that
          task.  Above-diagonal entries: compare performance across
          tasks (by subject).  Below-diagonal entries: empty?
          density/2d histograms?
    \begin{itemize}
    \item Memory-memory
    \item fitness-fitness
    \item survey-survey
    \item (fitness + survey)-memory
    \end{itemize}
  \item predictive analysis (regressions)
    \begin{itemize}
    \item Predict memory performance on held-out task from other tasks
    \item Predict memory performance on each task using fitness data
      \item Predict memory performance on each task using survey data
      \end{itemize}
    \item Reverse correlations: look at recent changes versus baseline
      trends (color using same scheme as behavior figures).  Possibly 
      \begin{itemize}
      \item Fitness profile that predicts performance on each task (barplots + timelines)
      \item Fitness profile for each survey demographic (barplots + timelines)
        \begin{itemize}
          \item Select out mental health demographics (based on meds, stress levels)
          \end{itemize}
        \end{itemize}
  \end{itemize}

>>>>

We used the same bootstrap procedure described above (used in our
exploratory correlation analyses) to generate $N = 1000$ bootstrapped
data matrices whose rows reflected sampled participants and whose
columns reflected different measured variables.  We used a round-robin
imputation procedure to estimate the values of any missing
features~\citep{Buck60}.  We applied this imputation procedure
independently for each bootstrapped data matrix to prevent data
contamination across iterations.

Next, we fit a series of ridge regression models to the training data.
We grouped variables according to whether they were memory-related,
fitness-related, or demographic-related.  
We used variables from each group to predict held-out variables from
the same group-- e.g., using performance on all but one task to
predict performance on the held-out task.  We also examined
predictions across groups of variables-- e.g., using fitness and
demographic variables to predict memory-related variables.

For each set of predictions, we used a bootstrap procedure to estimate
the stability and reliability of those predictions.  For each bootstrap
iteration, we divided the rows of that iteration's data matrix into
training and test sets.  The assignments of rows to these two sets was
random, subject to the constraint that any duplicated rows in the data
matrix (i.e., reflecting a single participant who had been sampled
multiple times) was always assigned to either the training \textit{or}
the test set--i.e., duplicated rows could not appear in both the
training and the test sets.  The training sets always comprised 75\%
of the data, and the tests sets comprised the remaining 25\% of the
data.

For each bootstrap iteration, we computed the root mean squared
deviation (RMSD) between the predicted and observed values in the
target features of the test dataset.  We defined this as the
\textit{observed} RMSD for that bootstrap iteration.  We also
estimated a null distribution of RMSD values by re-fitting each
iteration's data after randomly permuting the rows of the input data
matrix 
(thereby breaking any
meaningful associations between the input and output features).  Taken
together, this yielded a set of 1000 observed RMSD values and 1000
null RMSD values.  We assessed the statistical significance
($p$-values) of the observed RMSD values by computing the proportions
of null RMSD values that were less than the observed value.  We also
assessed the significance of the observed regression weights using
$t$-tests to compare the means of the observed versus null
distributions of weights, separately for each input and output feature.

@@

We wondered about potential links between the different aspects of
participants' data.  For example, if people who exercised in a
particular way also tended to perform better on a given memory task,
this could suggest that either (a) some property intrinsic to participants who
exercised in a particular way might also affect their memory
performance on the given task, and/or (b) particular exercise behaviors
could have a causal impact on memory performance.  We carried out an
exploratory analysis whereby we used a bootstrap-based approach (see
\textit{Exploratory correlation analyses}) to
identify reliable correlations between different aspects of memory
performance (Fig.~\behBehCorr), different aspects of fitness (Fig.~\fitFitCorr), different demographic
attributes (Fig.~\demDemCorr), and correlations between memory
performance, fitness information, and demographic attributes
(Fig.~\allCorr).  Several patterns emerged.  First, we found that
participants' performance on the (within-task) immediate versus delayed memory tests
from the free recall, naturalistic recall, and foreign language
flashcards tasks were positively correlated ($r$s $> 0.25$, $p$s $< 0.003$,
bootstrap corrected).  This suggests that, within each of these tasks,
similar processes or constraints may influence both short term and
long term information retrieval.  We also found reliable across-task correlations
between participants' (immediate and delayed) performance on the free
recall and foreign language flashcards tasks ($r$s $> 0.3$, $p$s $< 0.03$,
bootstrap corrected).

A large number of fitness-related measures displayed reliable
correlations (for a complete report, see Fig.~\fitFitCorr).  For
example, body mass index (BMI) and weight were correlated
($r = 0.91, p < 0.0001$, bootstrap corrected).  Resting heart rate over
the prior week was negatively correlated with recent
low-to-moderate-intensity (``fat burn'') cardiovascular
activity levels ($r = 0.70, p = 0.0004$, bootstrap corrected).
Participants' peak heart rate (averaged over the prior week) were also
negatively correlated with recent increases in step counts and daily
elevation gains ($r\mathrm{s} < -0.26, p\mathrm{s} < 0.03$, boostrap
corrected), where recent changes were defined as the average values
over the seven days leading up to the test day divided by the average
values over the preceding 30 days.  Although several demographic
attributes (Fig.~\demDemCorr) displayed trivial correlations (e.g.,
participants identifying as male never reported identifying as female,
and so on), we also observed a negative correlation between reported
stress and alertness ($r = -0.44, p < 0.0001$, bootstrap corrected), and
positive correlations between the reported clarity of the instructions for all
tasks ($r\mathrm{s} > 0.26, p\mathrm{s} < 0.02$, bootstrap corrected).

We also found reliable correlations between participants' fitness and
demographic measures and their behaviors in different tasks
(for a complete report, see Fig.~\allCorr).  For example, recent
low-to-moderate-intensity (``fat burn'') 
cardiovascular activity was positively
correlated with immediate ($r = 0.38, p = 0.03$) and delayed ($r =
0.38, p = 0.029$) recall on the naturalistic
memory task.  Recent increases in moderate-intensity (``cardio'')
activity over the prior 7 days (relative to the preceding 30 days) was
also positively correlated with immediate naturalistic recall ($r =
0.48, p = 0.003$) and immediate recall on the foreign language
flashcards task ($r = 0.43, p = 0.048$).  Recent high-intensity
(``peak'') activity was positively correlated with performance on the
spatial learning task ($r = 0.34, p < 0.0001$), as were recent increases
in high-intensity activity (prior 7 days versus preceding 30 days; $r
= 0.41, p = 0.01$).

\begin{itemize}
% \item characterizing behaviors (color by quartile and continue the
%   color scheme in later figures-- hue reflects task, shading reflects
%   performance.  white outline means immediate, black outline means delayed)
%   \begin{itemize}
%   \item Free recall (immediate + delayed): pfr, lag-CRP, spc  (color:
%     recall performance).  
%     Figure~\ref{fig:fr_behavioral}.
%     \item Naturalistic recall (immediate + delayed): reproduce a version of the sherlock
%       movie/recall trajectories (color: mean precision).  Figure~\ref{fig:nat_behavioral}.
%     \item Foreign language flashcards (immediate + delayed):
%       p(correct) histogram (color: p(correct)).  Figure~\ref{fig:vocab_behavioral}.
%       \item Spatial learning: mean error by number of shapes (color:
%         intercept and slope of line fit to errors as a function of the number of
%         shapes).  Figure~\ref{fig:spatial_behavioral}.
%       \end{itemize}
%     \item Fitness info (break down by task performance, potentially
%       separately for each task); also separate out recent (raw) and
%       recent versus baseline -- color using same color scheme as
%       behavior figure
%       \begin{itemize}
%       \item activity (steps, zone minutes, floors/elevation)
%       \item resting heart rate
%         \item sleep
%         \end{itemize}
  % \item exploratory analysis (correlations).  Possibly make some sort of scatter matrix or pairplot-- rows/columns:
  %         tasks.  Diagonal: histogram of performance metric for that
  %         task.  Above-diagonal entries: compare performance across
  %         tasks (by subject).  Below-diagonal entries: empty?
  %         density/2d histograms?
  %   \begin{itemize}
  %   \item Memory-memory
  %   \item fitness-fitness
  %   \item survey-survey
  %   \item (fitness + survey)-memory
  %   \end{itemize}
  \item predictive analysis (regressions)
    \begin{itemize}
    \item Predict memory performance on held-out task from other tasks
    \item Predict memory performance on each task using fitness data
      \item Predict memory performance on each task using survey data
      \end{itemize}
    \item Reverse correlations: look at recent changes versus baseline
      trends (color using same scheme as behavior figures).  Possibly 
      \begin{itemize}
      \item Fitness profile that predicts performance on each task (barplots + timelines)
      \item Fitness profile for each survey demographic (barplots + timelines)
        \begin{itemize}
          \item Select out mental health demographics (based on meds, stress levels)
          \end{itemize}
        \end{itemize}
  \end{itemize}
